---
title: "Microarray Data Analysis"
author: "Alireza Fathian"

output:
#  html_document:
#    toc: yes
  pdf_document:
    toc: yes
knit: (function(inputFile, encoding) {
  rmarkdown::render(inputFile, encoding = encoding, output_dir = "./../output") })
---
***
# Introduction
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Setting up project directory
```{r}
knitr::opts_knit$set(root.dir = '/home/alireza/Scripts/microarray_data_analysis')
```

# Import libraries
```{r, include=FALSE}
#install.packages(c("limma","GEOquery","Biobase","pheatmap",
#                   "ggplot2","gplots","reshape2","plyr"))
#install.packages("Biobase")
#install.packages("GEOquery")
#install.packages("limma")
#install.packages("gplots")
#install.packages("ggplot2")
#install.packages("plyr")
#install.packages("pheatmap")
```

```{r,message=FALSE}
library(limma)
library(Biobase)
library(GEOquery)
library(pheatmap)
library(ggplot2)
library(plyr)
library(pheatmap)
```

# Load Data

Importing GSE52509_series_matrix.txt.gz
```{r,message=FALSE,warning=FALSE,include=FALSE}
#gset = getGEO("GSE52509",GSEMatrix = TRUE, AnnotGPL = TRUE, destdir = "./data/raw")
```

```{r,message=FALSE, include=FALSE}
a = getwd()
strsplit(a, "/src", fixed = TRUE)
datadir = paste(strsplit(a, "/src", fixed = TRUE),"/data/raw/GSE52509_series_matrix.txt.gz",sep = "")
```

```{r,message=FALSE,warning=FALSE}
# import existing data
print(datadir)
gset=getGEO(filename=datadir, GSEMatrix = TRUE, AnnotGPL = TRUE)
```

# Dataset Description

```{r}
header=gset[[1]]
print(header)
```

Choosing shoerter column names:
```{r}
sml=c(rep("smoke_4",3),rep("control_4",3),rep("smoke_6",3),rep("control_6",3))
sml <- factor(sml)
levels(sml)
class(sml)
```

# Extracting Matrices
```{r}
ex<-exprs(gset)
class(ex)
dim(ex)
```

Frequency Histogram

```{r}
hist(ex[,1])
```

Normalizing Data

```{r}
ex<-log2(ex)
hist(ex[,1])
exprs(gset)<-ex
boxplot(ex,outline=FALSE)
```

If data was not normal

```{r}
x<-normalizeQuantiles(ex)
boxplot(x)
```

# Correlation HeatMap

```{r}
pheatmap(cor(ex))
pheatmap(cor(ex),labels_row = sml,labels_col = sml,legend = TRUE)
```


# Principal Component Analysis

```{r}
pca<-prcomp(ex)
plot(pca)
names(pca)
pca$sdev
colnames(pca$x)
plot(pca$x[,1:2])
```

```{r}
ex_scale=t(scale(t(ex),scale=F))
mean(ex_scale[1,])
pca2<-prcomp(ex_scale)
plot(pca2)
plot(pca2$x[,1:2])
```

```{r}
plot(pca2$r)
pc.sample<-data.frame(pca2$r[,1:3],Group=sml)
head(pc.sample)
ggplot(pc.sample,aes(PC1,PC2,color=Group))+geom_point(size=3)+theme_bw()
dev.off()
```

```{r}
pca2<-prcomp(t(ex_scale))
pc.sample<-data.frame(pca2$x[,1:3],Group=sml)
head(pc.sample)
ggplot(pc.sample,aes(PC1,PC2,color=Group))+geom_point(size=3)+theme_bw()
dev.off()

```

```{r}
sml <- factor(sml)
levels(sml)
sml
gset$description <- sml
design <- model.matrix(~ description + 0, gset) #112
colnames(design) <- levels(sml)
head(design)
design
```

# Differential Gene Expression

```{r}
fit <- lmFit(gset, design)
cont.matrix <- makeContrasts(smoke_4-control_4, levels=design)
fit2 <- contrasts.fit(fit, cont.matrix)
fit2 <- eBayes(fit2, 0.01)
```
Selecting first 250 genes with highest adjusted p-value
```{r}
tT <- topTable(fit2, adjust="fdr", sort.by="B", number=250)
tT <- topTable(fit2, adjust="fdr", sort.by="B", number=Inf)
colnames(tT)
head(ex)
tT <- subset(tT, select=c("ID","adj.P.Val","P.Value","t","B","logFC",
                          "Gene.symbol","Gene.title"))
tT <- topTable(fit2, adjust="fdr", sort.by="B", number=Inf)
colnames(tT)
head(ex)
tT <- subset(tT, select=c("ID","adj.P.Val","P.Value","t","B","logFC",
                          "Gene.symbol","Gene.title"))
```

```{r}
ms.up=subset(tT,select=c("ID","adj.P.Val","logFC","Gene.symbol"))
ms.up=subset(tT,tT$logFC>1 & tT$adj.P.Val<0.05, select=c( "ID", "adj.P.Val", "logFC", "Gene.symbol"))
```

```{r}
#write.table(ms.up.genenames,"/data/processed/up_ptA_ptB.txt", quote = F,row.names = F, col.names = F)
```
Finding out the up-regulated genes
```{r}
ms.up.genenames<- sub("///.*","",ms.up$Gene.symbol)
ms.up.genenames<- ms.up.genenames[ms.up.genenames!= ""]
ms.up.genenames<- strsplit2(ms.up.genenames,"///")
ms.up.genenames<- unique(ms.up.genenames)

ms.up.genenames
ms.up$Gene.symbol
```

```{r}
write.table(ms.up.genenames,"./data/processed/up_ptA_ptB.txt", quote = F,row.names = F, col.names = F)
```

